---
title: "Combined Data Cleaning and Exploratory Analysis Script for Salmonella Environmental Persistence Informs Management Relevant to Avian and Public Health"
author: "Kimberly M. Perez"
date: "10/28/2021"
output: html_document
---
# **Data Processing Methodology**
Before attention can be focused on the data analysis portion, removal of unnecessary variables and visualizing data will provide a clearer picture of the data. This section will focus on data cleaning and wrangling. First, both datasets must be loaded. Prior to loading any data into RStudio, ensure that a new Quarto or RMarkdown File is established and linked to the preferred working directory. Basic packages need to be loaded to read in the data. Such packages to download include read_csv(), read_excel(), and here(), each of which can be loaded by utilizing the library() function (e.g., library(here)). Next, the datasets from both the Ibis and Feeder trials are loaded into RStudio utilizing the appropriate function (e.g., read_csv() for .csv files and read_excel() for .xlsx files). After data is loaded into R, it is best to get acquainted with the data by utilizing basic functions such as glimpse(). Glimpse provides me with insight into my data such as column names, number of rows and how each row is classified (e.g., character or numeric). After viewing the data I can work on cleaning the data by removing unnecessary variables. I would like to make the data a little more uniform and easier to work with. With that, I will change "Yes" to 1, and *NAs* to 0. I believe this will help me further in the analysis process. Given the variablility in the column names, I will then clean the names up to make them easier to include in my code as I move forward. With all of that, it is finally time to visualize the data by making some tables and/or graphs. 

Here, some more packages were loaded that I forsaw using in this section. Before visualization began, the glimpse() function was utilized to view the recently cleaned data. Given the uncertainty of *Salmonella's* persistence on various surface materials and the information these trials hoped to provide, the summary() function was utilized to provide a quick glimpse of basic result summaries for positive *Salmonella* samples and if the pile present was the original pile or not. Next, I continued to explore these two variables of interest by plotting each using a stacked bar graph via ggplot().

I followed the same workflow for the Feeder Data, however, this dataset required less wrangling and did not provide as many exploratory options as the main focus was on feeder type, date, and the presence/absence of *Salmonella* for each day the feeders were sampled.

## **Data import and cleaning**

## **Initial Package Loading**
```{r}
# load a few R packages
library(here)
library(knitr)
library(dplyr)
library(readxl)
```

## **Picnic Table Data Loading**
```{r}
PTD<- read_excel(here("data", "raw_data", "Picnic_Table_Data.xlsx"))
View(PTD)
summary(PTD)
```
## **Picnic Table Data Wrangling**
```{r}
glimpse(PTD)

PTD1<- PTD %>% select(-c(Day, `Shippng Date`, Notes, `STAB TUBE NUMBER`, `ORIGIN OF FECES`, ...13)) #Removing columns

PTD1$`Original PP still intact?`<- ifelse(PTD1$`Original PP still intact?`=="Yes", 1, 0)

PTD2<- PTD1 %>% 
    mutate(`Original PP still intact?` = ifelse(is.na(`Original PP still intact?`), 0, `Original PP still intact?`)) #NAs to 0 for column `Original PP still intact?`

PTD3<- PTD2 %>% 
    mutate(`Pos orNeg`= ifelse(is.na(`Pos orNeg`), 0, `Pos orNeg`)) #NAs to 0 for column `Pos orNeg`

glimpse(PTD3) #Current variable names

PTD_clean<- PTD3 %>% #Renaming column names
  rename(
    ID= `Sample ID`,
    Sample_Type= `Sample Type (Pooled or Persistence)`,
    Salmonella_Positive= `Pos orNeg`,
    Original_Persistence_Pile= `Original PP still intact?`)

glimpse(PTD_clean)
```

## **Picnic Table Data Exploration and Visualization**
```{r}
#After cleaning I have decided I need to download a few more packages for the data visualization 

library(gtsummary) #Tables 
library(plotly) #I may try for an interactive plot, this is a good resource to use
library(ggplot2) #For basic visualization

PTDsumt1<- summary(PTD_clean$Salmonella_Positive) #Quick glimpse of samples, non-specific (e.g., persistence/pool) based on Salmonella positivity
PTDsumt1

PTDsumt2<- summary(PTD_clean$Original_Persistence_Pile) #Quick glimpse of samples, non-specific (e.g., persistence/pool) based on Persistence Piles (e.g., OG or not)
PTDsumt2

PTDH<- ggplot(PTD_clean, 
              aes(fill=Sample_Type, y= Salmonella_Positive, x= Date)) + 
  geom_bar (position="stack", stat="identity") +
  geom_vline(xintercept= as.numeric (PTD_clean$Date[c(95)]), linetype= "dotted", color="red", lwd=1)

PTDH #Basic Plot: Positive Salmonella Samples by Sample Type via stacked plot, red line denotes decrease, will tap Dr. Hernandez for more information on this decline (e.g., weather, shipping, etc.) 
ggsave(PTDH, file="Picnic_Table_PosSamp.png")

PTD_Quad<- ggplot(PTD_clean, 
              aes(fill=Quadrant, y= Salmonella_Positive, x= Date)) + 
  geom_bar (position="stack", stat="identity") +
  geom_vline(xintercept= as.numeric (PTD_clean$Date[c(95)]), linetype= "dotted", color="green", lwd=1)

PTD_Quad #Basic Plot: Positive Salmonella Samples (pooled v. persistence) by Quadrant via stacked plot green line denotes decrease, will tap Dr. Hernandez for more information on this decline (e.g., weather, shipping, etc.) 
ggsave(PTD_Quad, file="Picnic_Table_Quad.png")

```
## **Feeder Data Loading**
```{r}
library(readr)
FD<- read_csv(
  here("data", "raw_data", "Feeder Data.csv"))

View(FD)

summary(FD)

str(FD)

glimpse(FD)
```

## **Feeder Data Wrangling**
```{r}
FD1<- FD %>% select(-c(`Absence (0)/Presence (1)`, `...5`)) #Removing columns

glimpse(FD1) #Quick glimpse at the data, I can determine if I should wrangle the data a bit more

FD_clean<- FD1 #Data looked clean so I will rename it to use with my data exploration and visualization process below 
```

**Picnic Table Data Exploration and Visualization**
```{r}
summary(FD_clean$Absence_0_Presence_1) # Utilizing summary() for a quick look at result summaries- not much info here

FD_plot<- ggplot(FD_clean, #Given my limited dataset I will focus my visualization on feeder type and presence/absence of Salmonella 
              aes(fill=Feeder_Type, y=Absence_0_Presence_1, x= Collection_Date)) + 
  geom_bar (position="stack", stat="identity") 

FD_plot  
ggsave(FD_plot, file="Feeder_PA.png")
```