---
title: "Supplement to *Salmonella* Environmental Persistence Informs Management Relevant to Avian and Public Health for a Data Analysis Project"
author: Kimberly Perez
date: "`r Sys.Date()`"
output: html_document
---

# **Overview**
The information included on this .qmd file is code that was written but not utilized in final version of my manuscript. The majority of what you will find on this page comes directly from my "StatisticalAnalysis.qmd".  

Loading in the packages we may need...
```{r}
#Load needed packages
library(tidyverse)
library(ggplot2) #For plotting/data vis.
library(broom) #For cleaning up output from lm()
library(dotwhisker) #Visualization for regression outputs
library(here) #Setting paths/for data loading/saving
library(tidymodels) #Great package for models and other resources used for fitting
library(tidyr) #Helps with data wrangling
library(gtsummary) #To create tables 
library(dplyr)
library(vip)
library(performance) #Model evaluation
library(jtools)
library(broom)
library(broom.mixed) #Converts fitted to tiddy dfs
library(finalfit) #Needed for Time to Event Analysis 
library(forcats) #Needed for Time to Event Analysis 
library(survival) #Needed for Time to Event Analysis 
library(lubridate)
library(ggsurvfit)
library(gt)
library(knitr) #For tables
```

We need the data from both trials, so let's load them in using here() 
```{r}
#Picnic Table Data
PT_Data<- readRDS(here::here("2. Clean Data","PTprocessed.rds"))

#Feeder Data 
FD_Data <-readRDS(here::here("2. Clean Data","FDprocessed.rds"))

glimpse(PT_Data)

glimpse (FD_Data)
```


```{r}
#### Model fit another way
# fit linear model using Salmonella_Pos as outcome, Table as predictor
lmfit3.0 <- lm(Salmonella_Positive~Table, PT_Data) 
lmfit3.0
plot(lmfit3.0)

# Here we place results from the above fit into a data frame via tidy function
lmtable1 <- broom::tidy(lmfit3.0)

#Let's check out the results
print(lmtable1)

#A different way 
# I will do the same as above, however, I am now adding persistence pile
lmfit2 <- lm(Salmonella_Positive~Original_Persistence_Pile+Table, PT_Data) 
plot(lmfit2)

# Convert to a data frame via tidy function
lmtable2 <- broom::tidy(lmfit2)

#Results
print(lmtable2)
```


This section is a continuation of what was produced above, however, we will now be utilizing the Feeder Trial Data

```{r}
#Doing the same for FEEDER DATA
lmog_FD<-linear_reg() %>% 
  set_engine("lm")

model_lmFD<- linear_reg()

lmfit_FD<- model_lmFD %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Absence_0_Presence_1~Feeder_number, FD_Data) 

lmfit_FD #Let's view what we just did

glimpse(lmfit_FD)

tidy(lmfit_FD) #We can use the tidy() function to produce the output/ summary statistics

plot_summs(lmfit_FD)

#Checking model performance: FEEDER
check_model(lmfit_FD$fit)
```

**Time to Event Analysis: Picnic Table Trial**
```{r}

PT_Data_TTE<- #First let's format the dates 
  PT_Data %>% 
  mutate(
    Date= ymd(Date))

PT_Data_TTE$Date<- as.numeric(PT_Data_TTE$Date) #Let's make these numeric

PT_Data_TTE <- #I want to add a column displaying time between events (by day)
  PT_Data_TTE %>%
  mutate(
    os_day= as.duration(Date)/ddays(1))

PT_Data_TTE$os_day<- as.numeric(PT_Data_TTE$os_day) #Again let's make sure our new column is in the proper format

surv1<- Surv(PT_Data_TTE$Date,PT_Data_TTE$Salmonella_Positive) 

#Let's visualize!
sfit1<- survfit(Surv(Date, Salmonella_Positive) ~  1, data=PT_Data_TTE)  %>% 
  ggsurvfit() + 
  labs(
    x = "Days",
    y = "Overall Salmonella Persistence") + #Let's add some labels
  add_confidence_interval() +
  add_risktable()

#I am not confident with this graphic so I will include this in my supplemental section
```

```{r}
glimpse(PT_Data)


PT_Data = PT_Data %>% 
 mutate(
  salmonella_crr= case_when(
   Salmonella_Positive ==  0~1,
    Salmonella_Positive == 1~1,
     TRUE ~ 1))

dependent_crr= "Surv(Date, salmonella_crr)"
explanatory= c("Table", "Quadrant", "Sample_Type")

PT_Data$Date<- as.numeric(PT_Data$Date)

PT_Data %>% finalfit(dependent_crr, explanatory)

PT_Data %>% 
   finalfit(dependent_crr, explanatory, add_dependent_label = FALSE) %>% 
    rename("Overall survival" = label) %>% 
    rename(" " = levels) %>% 
    rename("  " = all) 

survfit(Surv(Date, Salmonella_Positive)~1, data=PT_Data)

PT_MedSurv<-PT_Data %>%
 filter(salmonella_crr==1) %>%
  summarize(median_surv=median(Date))

PT_MedSurv <- ((PT_MedSurv)/31556952000)*100

PT_Surv_Table<- gt(PT_MedSurv) %>%
   tab_header(
    title = "Median Survival of Salmonella",
    subtitle = ("01/02/2020 to 01/09/2020")) %>%
  cols_label(
    median_surv= "Median Salmonella Survival (days)") 

PT_Surv_Table # persistence seems to be roughly 5 days
```

**Time to Event Analysis: Feeder Trial**
```{r}
FD_Data_TTE<- #First let's format the dates 
  FD_Data %>% 
  mutate(
    Date= ymd(Collection_Date))

FD_Data_TTE$Collection_Date<- as.numeric(FD_Data_TTE$Collection_Date) #Let's make these numeric

FD_Data_TTE <- #I want to add a column displaying time between events (by day)
  FD_Data_TTE %>%
  mutate(
    os_day= as.duration(Collection_Date)/ddays(1))

FD_Data_TTE$os_day<- as.numeric(FD_Data_TTE$os_day) #Again let's make sure our new column is in the proper format

surv1_FD<- Surv(FD_Data_TTE$Collection_Date,FD_Data_TTE$Absence_0_Presence_1) 

#Let's visualize!
sfit1<- survfit(Surv(Collection_Date, Absence_0_Presence_1) ~  1, data=FD_Data_TTE)  %>% 
  ggsurvfit() + 
  labs(
    x = "Days",
    y = "Overall Salmonella Persistence") + #Let's add some labels
  add_confidence_interval() +
  add_risktable()

sfit1
```

```{r}
FD_Data = FD_Data %>% 
 mutate(
  salmonella_crr= case_when(
   Absence_0_Presence_1 ==  0~1,
    Absence_0_Presence_1 == 1~1,
     TRUE ~ 1))

dependent_crr= "Surv(Collection_Date, salmonella_crr)"
explanatory= c("Feeder_Type")

FD_Data$Collection_Date<- as.numeric(FD_Data$Collection_Date)

FD_Data %>% finalfit(dependent_crr, explanatory)

FD_Data %>% 
   finalfit(dependent_crr, explanatory, add_dependent_label = FALSE) %>% 
    rename("Overall survival" = label) %>% 
    rename(" " = levels) %>% 
    rename("  " = all) 

survfit(Surv(Collection_Date, Absence_0_Presence_1)~1, data=FD_Data)

FD_MedSurv<-FD_Data %>%
 filter(salmonella_crr==1) %>%
  summarize(median_surv=median(Collection_Date))

FD_MedSurv <- ((FD_MedSurv)/604800)*100 #Persistence seems to be roughly 3 days for the FD

FD_Surv_Table<- gt(FD_MedSurv) %>%
   tab_header(
    title = "Median Survival of Salmonella",
    subtitle = ("02/22/2022 to 03/14/2022")) %>%
  cols_label(
    median_surv= "Median Salmonella Survival (days)") 

FD_Surv_Table # seems to be roughly 3 days
```

**Cox regression: Picnic Table Trial**

```{r}
PT_Data$Date<- as.numeric(PT_Data$Date)

coxph(Surv(Date, Salmonella_Positive) ~ Sample_Type, PT_Data)

CR_PT1<-coxph(Surv(Date, Salmonella_Positive) ~ Sample_Type, PT_Data) %>%
  tbl_regression(exp=TRUE)

CR_PT1
```

**Cox regression: Feeder Trial**

```{r}
FD_Data$Collection_Date<- as.numeric(FD_Data$Collection_Date)

coxph(Surv(Collection_Date, Absence_0_Presence_1) ~ Feeder_Type, FD_Data)

CR_FD1<- coxph(Surv(Collection_Date, Absence_0_Presence_1) ~ Feeder_Type, FD_Data) %>%
  tbl_regression(exp=TRUE)

CR_FD1

CR_FD1%>%
  saveRDS(here("4. Results", "FD_Cox.rds"))
```